Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-09-25 paper/sections/proof-reconstruction.tex
[ paper ] reordering formalisation added.

\emph{Splitting a goal.}
The goal of a problem to be proved by \Metis is processed to generate new
propositions, the new targets, \Metis calls these formulas
\emph{subgoals}.

@@

% To deal with nested conjunctions and nested disjunctions, we follow a similar strategy mainly described in \citeauthor{bohme2010}~\cite{bohme2010}.

First, we define the $rconj$ function in Eq.~\ref{eq:rconj-def} to convert conjunctions to be right-associative conjunctions, and in a similar way with disjunctions,
we define $rdisj$ function, to convert a general disjunction to a be right-associative disjunction.

\begin{equation}
\label{eq:rconj-def}
\begin{split}
&rconj(φ) =\begin{cases}
ψ₁ ∧ rconj(ψ₂ ∧ φ₂) &\text{if }φ = φ₁ ∧ φ₂\text{, } ψ≡rconj(φ₁),\\
&\text{and }ψ ≡ ψ₁ ∧ ψ₂,\\[3mm]
ψ ∧ rconj(φ₂)  &\text{if }φ = φ₁ ∧ φ₂\text{, } ψ≡rconj(φ₁),\\
φ &\text{otherwise.}
\end{cases}
\end{split}
\end{equation}

\begin{equation}
\begin{split}
\label{eq:buildOr}
buildOr(φ, ψ) =
\begin{cases}
ψ &\text{if } ψ ≡ ψ₁ ∨ ψ₂\text{ and } buildOr(φ, ψ_{i}) ≡ ψ_{i}\text{ for some }i=1, 2,\\
ψ &\text{if } φ ≡ ψ,\\
φ &\text{otherwise.}
\end{cases}
\end{split}
\end{equation}

>>>>

\subsubsection{Splitting a Goal.}
The goal of a problem to be proved by \Metis is processed to generate new
propositions, the new targets, \Metis calls these formulas
\emph{subgoals}.

@@

% ------------------------------------------------------------------------

\subsubsection{Reordering.}

Syntactically speaking, showing the logic equivalence between two conjunctive normal forms is a matter of rearrangements of the conjunctions
but also of the inner disjunctions inside the propositions involved.

We provide some useful theorems to perform the following tasks.

\begin{itemize}
  \item Reorder a disjunction like $φ = φ₁ ∨ φ₂ ∨ \cdots$ to match with another disjunction $ψ = ψ₁ ∨ ψ₂ ∨ \cdots$, when $φ$ is a disjunct of $ψ$.
  \item Reorder a conjunction of disjuncts (CNF), $φ = φ₁ ∧ φ₂ ∧ \cdots$, to match with another CNF formula equivalent $ψ = ψ₁ ∧ ψ₂ ∧ \cdots$.
\end{itemize}

First, we define the $rdisj$ function in Eq.~\ref{eq:rdisj-def} to convert a disjunction to be a right-associative disjunction, and following
a similar approach we define the $rconj$ function for conjunctions.

\begin{align}
    \begin{split}
      \label{eq:rdisj-def}
      rdisj(φ) =
      \begin{cases}
      rdisj(φ₁ ∨ (φ₂ ∨ φ₃)) & \text{if } φ ≡ (φ₁ ∨ φ₂) ∨ φ₃, \\
      φ₁ ∨ rdisj(φ₂)        & \text{if } φ ≡ φ₁ ∨ φ₂, \\
      φ                     & \text{otherwise.}
      \end{cases}
    \end{split}
\end{align}

\begin{lemma}[lem-rdisj]
\label{lem:lem-rdisj}
  $Γ ⊢ φ$ then $Γ ⊢ rdisj(φ)$.
\end{lemma}

The following function, $build_{∨} : \Prop → \Prop → \Prop$ in Eq.~\ref{eq:build-or}, intends to construct a disjunction from another disjunction, specifically, we want to rearrange the order of the disjuncts in a target disjunction.
This approach for rearrangement is inspired on the treatment for nested
disjunctions mainly described in \citeauthor{bohme2010}~\cite{bohme2010}.

\begin{equation}
\begin{split}
\label{eq:build-or}
&build_{∨}(φ, ψ) =\\
&\hspace{3mm}
\begin{cases}
ψ &\text{if } φ ≡ ψ,\\
ψ &\text{if } ψ ≡ ψ₁ ∨ ψ₂\text{ and } build_{∨}(φ, ψ_{i}) ≡ ψ_{i}\text{ for some }i=1, 2,\\
φ &\text{otherwise.}
\end{cases}
\end{split}
\end{equation}

\begin{lemma}[lem-build-or]
\label{lem:lem-build-or}
$Γ ⊢ φ$, and $ψ\in \Prop$ then $Γ ⊢ build_{∨}(φ)$.
\end{lemma}

The function $factor : \Prop → \Prop$  in Eq.~\ref{eq:factor-definition} is intended to work only for right-associative disjunctions, and reduces a disjunction of the same disjunct (e.g., $factor(φ ∨ (φ ∨ \cdots)) \mapsto\ φ$).

\begin{align}
\begin{split}
  \label{eq:factor-definition}
  factor(φ) =
  \begin{cases}
    φ₁ & \text{if }φ ≡ φ₁ ∨ φ₂ \text{ and } φ₁ ≡ factor(φ₂),\\
    φ  & \text{otherwise.}
  \end{cases}
\end{split}
\end{align}

\begin{lemma}[lem-factor]
\label{lem:lem-factor}
$Γ ⊢ φ$ then $Γ ⊢ factor(φ)$.
\end{lemma}

Now, with a right-associative proposition,
$strongBuild_{∨}: \Prop → \Prop → \Prop$ in
Eq.~\ref{eq:strong-build-or-definition} aims to construct the disjunction
ψ by using each of disjuncts from the proposition φ. This approach to
build a disjunction from another is inspired on the description
\citeauthor{bohme2010}~\cite{bohme2010} to prove equality between nested
disjunctions.

\begin{align}
  \begin{split}
  \label{eq:strong-build-or-definition}
    &strongBuild_{∨}(φ, ψ) =\\
    &\hspace{3mm}
    \begin{cases}
      factor(build_{∨}(φ₁, ψ) ∨ build_{∨}(φ₂,ψ))  & φ≡ φ₁ ∨ φ₂,\\
      build_{∨}(φ, ψ)  & \text{otherwise.}
    \end{cases}
  \end{split}
\end{align}

\begin{lemma}[lem-sbuild-or]
\label{lem:lem-sbuild-or}
$Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ strongBuild_{∨}(φ, ψ)$.
\end{lemma}

Let see the following example of the idea behind $strongBuild_{∨}$
function; from the disjunction $(r ∨ (q ∨ p))$, we build the disjunction
$((p ∨ q) ∨ r)$.

\begin{equation*}
  \begin{bprooftree}
  \AxiomC{Γ ⊢ q}
  \RightLabel{∨-intro₂}
  \UnaryInfC{Γ ⊢ p ∨ q}
  \RightLabel{∨-intro₁}
  \UnaryInfC{Γ ⊢ (p ∨ q) ∨ r}

  \AxiomC{Γ ⊢ p}
  \RightLabel{∨-intro₁}
  \UnaryInfC{Γ ⊢ p ∨ q}
  \RightLabel{∨-intro₁}
  \UnaryInfC{Γ ⊢ (p ∨ q) ∨ r}

  \RightLabel{∨-elim}
  \BinaryInfC{Γ, q ∨ p ⊢ (p ∨ q) ∨ r}
  \end{bprooftree}
\end{equation*}

\begin{equation*}
  \begin{bprooftree}
  \AxiomC{Γ ⊢ r}
  \RightLabel{∨-intro₂}
  \UnaryInfC{Γ ⊢ (p ∨ q) ∨ r}

  \AxiomC{Γ, q ∨ p ⊢ (p ∨ q) ∨ r}

  \RightLabel{∨-elim}
  \BinaryInfC{Γ, r ∨ (q ∨ p) ⊢ (p ∨ q) ∨ r}

  \RightLabel{⇒-intro}
  \UnaryInfC{Γ ⊢ r ∨ (q ∨ p) ⇒ (p ∨ q) ∨ r}

  \end{bprooftree}
\end{equation*}

\begin{theorem}[thm-reorder$_{∨}$] $Γ ⊢ φ$, and $ψ\in\Prop$ then $Γ ⊢ reorder_{∨}(φ, ψ)$, where,
\begin{equation*}
  \label{eq:reorder-or}
  reorder_{∨} (φ, ψ) = strongBuild_{∨}(rdisj(φ), ψ).
\end{equation*}
\end{theorem}
\begin{proof} Use Lemma~\ref{lem:lem-rdisj} and Lemma~\ref{lem:lem-sbuild-or}.
\end{proof}

To work with nest conjunctions of disjunctions (CNFs), we define similar
functions and theorems as we previously did for disjunctions.

\begin{lemma}[lem-reorder$_{∧}$]
\label{lem:lem-reorder-and}
$Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ reorder(φ, ψ)$, where,
\begin{align*}
    \begin{split}
      reorder_{∧}(φ, ψ) =
      \begin{cases}
        φ &\text{if }φ ≡ ψ, \\
        ψ₁ ∧ ψ₂ &\text{if } ψ ≡ ψ₁ ∧ ψ₂\text{, }reorder_{∧}(φ, ψ₁) ≡ ψ₁,\\
                &\text{and }reorder_{∧}(φ, ψ₂)≡ ψ₂,\\
        φ       &\text{if } ψ ≡ ψ₁ ∧ ψ₂,\\
        conjunct(φ, ψ) &\text{otherwise.}
      \end{cases}
    \end{split}
\end{align*}

\end{lemma}

\begin{lemma}[lem-conjunct$_{∨}$]
\label{lem:lem-conjunct-or}
$Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ conjunct_{∨}(φ, ψ)$, where,

\begin{align*}
\begin{split}
  \label{eq:conjunct-or-definition}
  conjunct_{∨}(φ, ψ)=
  \begin{cases}
    ψ &\text{if }reorder_{∨}(φ, ψ) ≡ ψ,\\
    ψ &\text{if }ψ ≡ ψ₁ ∧ ψ₂, conjunct_{∨}(φ, ψ₁) = ψ₁,\\
      &\text{and } reorder_{∨}(φ, ψ₂) ≡ ψ₂,\\
    ψ &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₁, ψ) ≡ ψ,\\
    ψ &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₂, ψ) ≡ ψ,\\
    φ &\text{otherwise.}
  \end{cases}
\end{split}
\end{align*}

\end{lemma}

\begin{theorem}[thm-reorder$_{∧∨}$]
\label{thm:thm-reorder-or-and}
$Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ reorder_{∧∨}(φ, ψ)$, where,

\begin{align}
    \begin{split}
    reorder_{∧∨}(φ, ψ)=
      \begin{cases}
        ψ &\text{if } φ≡ψ,\\
        ψ &\text{if } ψ ≡ ψ₁ ∧ ψ₂, reorder_{∧∨}(φ,ψ₁) ≡ ψ₁,\\
          &\text{and }reorder_{∧∨}(φ, ψ₂) ≡ ψ₂,\\
        φ &\text{if }ψ ≡ ψ₁ ∧ ψ₂,\\
        conjunct_{∨}(φ, ψ) &\text{otherwise.}
      \end{cases}
    \end{split}
\end{align}

\end{theorem}
